---
title: "Assignment 6: GLMs week 1 (t-test and ANOVA)"
author: "Nikki Shintaku"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on t-tests and ANOVAs. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 18 at 1:00 pm.

## Set up your session

1. Check your working directory, load the `tidyverse`, `cowplot`, and `agricolae` packages, and import the NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv dataset. 

2. Change the date column to a date format. Call up `head` of this column to verify.

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(cowplot)
library(agricolae)
library(ggthemes)

lake_nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

#2
lake_nutrients$sampledate <- as.Date(lake_nutrients$sampledate , format = "%Y-%m-%d")
head(lake_nutrients$sampledate)

```


## Wrangle your data

3. Wrangle your dataset so that it contains only surface depths and only the years 1993-1996, inclusive. Set month as a factor.
```{r}
#Wrangle the data
lake_nutrients_depth <- 
  lake_nutrients %>%
  filter(year4 %in% c("1993", "1994", "1995", "1996")) %>%
  filter(depth_id == 1)

#change to month to factor
lake_nutrients_depth$month <- as.factor(lake_nutrients_depth$month)

```

## Analysis

Peter Lake was manipulated with additions of nitrogen and phosphorus over the years 1993-1996 in an effort to assess the impacts of eutrophication in lakes. You are tasked with finding out if nutrients are significantly higher in Peter Lake than Paul Lake, and if these potential differences in nutrients vary seasonally (use month as a factor to represent seasonality). Run two separate tests for TN and TP.

4. Which application of the GLM will you use (t-test, one-way ANOVA, two-way ANOVA with main effects, or two-way ANOVA with interaction effects)? Justify your choice.

> Answer: I choose to run a Two-Way ANOVA with interaction effects. An ANOVA is being used because there is three or more groups within the seasons (months) so a T-Test would not be appropriate. A two-way ANOVA is being used because we want to examine the effects of two categorical explanatory variables on a continuous varible. Lake and Month are the two categorical variables being tested on total nitrogen and total phosphorus which are continous variables. Interaction effects is included in the two-way ANOVA because the effects of lake and month may be dependent on each other especially since Peter Lake was manipulated with nutrients over our time period. A two-way ANOVA with interaction effects will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

5. Run your test for TN. Include examination of groupings and consider interaction effects, if relevant.

6. Run your test for TP. Include examination of groupings and consider interaction effects, if relevant.

```{r}
#5
TN_interaction <- aov(data = lake_nutrients_depth, tn_ug ~ month * lakename)
summary(TN_interaction)
#There is no significant interaction effect on month:lakename; P value = 0.379

TN_interaction_lm <- lm(data = lake_nutrients_depth, tn_ug ~ month * lakename)
summary(TN_interaction_lm)
#Peter lake has higher total nitrogen than Paul Lake and do not vary seasonally 
#no further tests need to be run 

#6
TP_interaction <- aov(data = lake_nutrients_depth, tp_ug ~ month * lakename)
summary(TP_interaction)
#There is a significant interaction effect --> interpret interaction effects only

TP_interaction_lm <- lm(data = lake_nutrients_depth, tp_ug ~ month * lakename)
summary(TP_interaction_lm)
#Peter lake has higher total phosphorus than Paul Lake

#Run a post-hoc test for pairwise differences
TukeyHSD(TP_interaction)

TP_interaction.lake.month <- with(lake_nutrients_depth, 
                                  interaction(month, lakename))
TP_interaction.lake.month.aov <- aov(data = lake_nutrients_depth, 
                                     tp_ug ~ TP_interaction.lake.month)

TP_interaction_groups <- HSD.test(TP_interaction.lake.month.aov, 
                                  "TP_interaction.lake.month", group = TRUE)
TP_interaction_groups
```

7. Create two plots, with TN (plot 1) or TP (plot 2) as the response variable and month and lake as the predictor variables. Hint: you may use some of the code you used for your visualization assignment. Assign groupings with letters, as determined from your tests. Adjust your axes, aesthetics, and color palettes in accordance with best data visualization practices.

8. Combine your plots with cowplot, with a common legend at the top and the two graphs stacked vertically. Your x axes should be formatted with the same breaks, such that you can remove the title and text of the top legend and retain just the bottom legend. 

```{r}
mytheme <- theme_stata(base_size = 14, base_family = "sans", scheme = "s2mono") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)

#7
#Total Nitrogen Plot
TN_plot <- ggplot(lake_nutrients_depth, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(x = "Month", y = expression(paste("Total Nitrogen (", mu, "g/L)")), 
       color = "Lake Name") +
  scale_color_stata("s2color") +
  ylim(0,2100)

print(TN_plot)

#Total Phosphorus Plot
TP_plot <- ggplot(lake_nutrients_depth, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(x = "Month", y = expression(paste("Total Phosphorus(", mu, "g/L)")), 
       color = "Lake Name") +
  scale_color_stata("s2color") +
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("cd", "bcd", "d", "ab", "d", "a", 
                         "d", "abc", "cd", "abcd"),
               show.legend = FALSE,
               position = position_dodge2(0.6)) +
  ylim(0,70)

print(TP_plot)

#8 
boxplots_nutrients <- 
  plot_grid(TN_plot + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank()),
            TP_plot + theme(legend.position = "none"),
            labels = "AUTO", 
            align = "h", 
            axis = "b",
            ncol = 1)
print(boxplots_nutrients)
```




